Lattice Boltzmann Microchannel Simulations via Binary Liquid Model

### 2011-05-06 Conference/CFD2011/paper.tex
Finished Jos revision. Waiting for the feedback. Switching to rectangular channels. High performance issues and mass transfer.

\title{Lattice Boltzmann Microchannel Simulations via Binary Liquid Model}
\paperID{CFD11-52}
\author{Alexandr}{Kuzmin} %{forename}{surname}
\presenting  %the previous author is presenting the paper (name becomes underlined)
\email{kuzmin@ualberta.ca}%e-mail address of the previous author
\address{Chemical and Materials Engineering, University of Alberta, 9107-116 Street,  T6G 2V4
Edmonton, CANADA}%affiliation of the previous author
\author{Dmitry}{Eskin}
\email{deskin@slb.com}
\address{DBR Schlumberger Technology Center, 9450 17 Ave NW, T6N 1M9 Edmonton,CANADA}
\author{Jos}{Derksen}
%\address{NTNU Department of Physics, 7491 Trondheim, CANADA}
\email{jos@ualberta.ca}
\address{Chemical and Materials Engineering, University of Alberta, 9107-116 Street,  T6G 2V4
Edmonton, CANADA}%affiliation of the previous author

@@

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} also used the finite volume method to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. 
\citet{heil-bretherton} and \citet{ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using the finite element method. \citet{giavedoni-numerical} performed cross validation of
the
finite element solution with previously published results.
The solutions were obtained for circular and planar cases. 
%The planar case results for the film
%thickness dependency on the capillary number are indicated in Fig. \ref{fig:giavedoni:planar}.
%\begin{figure}[ht]
%\includegraphics[width=0.5\textwidth]{Figures/giavedoni_planar.eps}
%\caption{\citet{giavedoni-numerical} gathered results across the
%literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
%scaled to the half-width of the
%channel. $h_{\infty}$ is the film thickness of gas finger at
%the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
%\end{figure}

@@

The experimental works supported by numerical simulations reveal interesting phenomena happening in
three-dimensional geometry cases. It was found \cite{heil-threedim,wong-films} that for rectangular
or square shaped capillaries there is a transition for a
certain capillary number, where the bubble crosssection changes from the non axisymmetric to axis
symmetric case,
Fig. \ref{fig:sym:antisym}.
From hereon, we limit ourselves to the case of microchannels with square crosssections.
Non axisymmetric case is attributed to the case where the axial radius is different from the
diagonal radius and the bubble has the non-circular shape in the channel crossection. In this case
the bubble shape mimics the shape of the square and looks like a rounded square. The dependence
of the diagonal and axial radii on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the deposited liquid for the range of
capillary number. One can see the asymmetry between diagonal and axis diameter for the capillary
number $Ca\leq\hat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The transition number between non-axisymmetric case and symmetric is indicated in a number of
works as $\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space}, $\widehat{Ca}=0.033$ \cite{heil-threedim}. If the capillary number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the droplet dependant on the capillary number. One of the example for the bubble radii
dependence on the capillary number is in Fig. \ref{fig:heil:three:dim}.

@@

circular shaped capillaries. They also found an empirical correlation which allows to put

@@

The lattice Boltzmann method has emerged as a successful method to simulate

@@

\section{Numerical benchmark}
To be able to properly simulate and compare simulation results with data
published in the literature, one needs to design a numerical benchmark addressing
all the challenges for the lattice Boltzmann method. That includes the diffuse nature of the
interface which needs to have a negligible effect on the results. Also, as far as the density is
uniform, the binary liquid model can address only different viscosities of gas and liquid.
Another challenge is boundary conditions because one needs to couple inlet and outlet
boundary conditions \cite{heil-threedim} for the film thickness to establish.  The geometry of
the problem is also of concern. Most literature numerical simulations are done for a channel with
circular cross section,
which is quite difficult to address in terms of lattice Boltzmann framework. We thoroughly discuss
all the
challenges associated with the lattice Boltzmann microchannel simulations in our recent work
\cite{kuzmin-binary2d}.

@@

Based on these results one can conclude
that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.07$.
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$. $\delta$ is scaled on $H_{\mathrm{eff}}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number we aim for is $\Ca=0.05$. The capillary number obtained from simulations
$\Ca=0.07$. 
\label{fig:grid:profiles}}
\end{figure}
\begin{table}
\begin{tabularx}{0.5\textwidth}{|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$\\
\hline
$1500$&$102$&$0.0694$&$0.0041$&$0.824$&$200000$\\
\hline
$1875$&$127$&$0.0688$&$0.0041$&$0.646$&$250000$\\
\hline
$2250$&$152$&$0.0676$&$0.0040$&$0.539$&$300000$\\
\hline
$2625$&$177$&$0.0679$&$0.0040$&$0.453$&$350000$\\
\hline
$3000$&$202$&$0.0668$&$0.0041$&$0.400$&$400000$\\
\hline
$3375$&$227$&$0.0663$&$0.0039$&$0.355$&$450000$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the interface velocity of the bubble center
measured at the front meniscus.  $5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta
(N_y-2)$ is the size of the film in
lattice Boltzmann units.
\label{table:parameters:grid:refinement}}
\end{table}

@@

\section{Conclusion}
This work presents numerical studies of the
Bretherton/Taylor problem using the binary liquid lattice Boltzmann method. The bubble was chosen
sufficiently long for the film
thickness to
stabilize, and periodic boundary conditions were used to keep the simulations robust. The
distance between bubbles was taken large enough to minimize
the mutual influence of neighboring bubbles. The computational
results in terms of capillary number dependence and shape of the bubbles show consistency with the
previously published data in two and three dimensions. The simulations show that in the range of low
capillary numbers the binary liquid model can be used to
simulate gas finger/bubbles propagation in the microchannel. The model is able to catch not only
the film widths but also the velocity patterns change. 
 An examination of the influence of grid resolution on the
results allowed
us to determine that the phase interface should be resolved as at least $50$ percent of the film
thickness
in order for the simulations to be grid independent. Though our results are specific to the binary
liquid lattice
Boltzmann method, the numerical hints and procedures can be used for any
continuous interface method.

>>>>

\title{Lattice Boltzmann Microchannel Simulations via Binary Liquid Model}
\paperID{CFD11-94}
\author{Alexandr}{Kuzmin} %{forename}{surname}
\presenting  %the previous author is presenting the paper (name becomes underlined)
\email{kuzmin@ualberta.ca}%e-mail address of the previous author
\address{Chemical and Materials Engineering, University of Alberta, 9107-116 Street,  T6G 2V4
Edmonton, CANADA}%affiliation of the previous author
\author{Dmitry}{Eskin}
\email{deskin@slb.com}
\address{DBR Schlumberger Technology Center, 9450 17 Ave NW, T6N 1M9 Edmonton,CANADA}
\author{Jos}{Derksen}
%\address{NTNU Department of Physics, 7491 Trondheim, CANADA}
\email{jos@ualberta.ca}
\address{Chemical and Materials Engineering, University of Alberta, 9107-116 Street,  T6G 2V4
Edmonton, CANADA}%affiliation of the previous author

@@

There are a number of numerical methods which
were used for the simulation of the Taylor/Bretherton flow.
\citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillary using the finite volume method.
\citet{kreutzer-pressure-drop} also used the finite volume method to perform
simulations of a circular capillary for a number of different
Reynolds and capillary numbers. 
\citet{heil-bretherton} and \citet{ingham-plates} studied gas finger propagation in
a two-dimensional channel for a range of Reynolds and capillary
numbers using the finite element method. \citet{giavedoni-numerical} performed cross validation of
the
finite element solution with previously published results.
The solutions were obtained for circular and planar cases. Smooth particle hydrodynamics (SPH) is
presented by \citet{huang-sph}. However except showing that the SPH is able to reproduce the
gas-liquid interface, there is no thorough study of the deposited film thickness. 
%The planar case results for the film
%thickness dependency on the capillary number are indicated in Fig. \ref{fig:giavedoni:planar}.
%\begin{figure}[ht]
%\includegraphics[width=0.5\textwidth]{Figures/giavedoni_planar.eps}
%\caption{\citet{giavedoni-numerical} gathered results across the
%literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
%scaled to the half-width of the
%channel. $h_{\infty}$ is the film thickness of gas finger at
%the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
%\end{figure}

@@

The experimental works supported by numerical simulations reveal interesting phenomena occuring in
three-dimensional geometry cases. It was found \cite{heil-threedim,wong-films} that for rectangular
or square shaped capillaries there is a transition for a
certain capillary number, where the bubble crosssection changes from the non-axisymmetric to
axisymmetric case,
Fig. \ref{fig:sym:antisym}.
From hereon, we limit ourselves to the case of microchannels with square crosssections.
Non-axisymmetric case is attributed to the situation where the axial radius is different from the
diagonal radius and the bubble has the non-circular shape in the channel crossection. In this case
the bubble shape mimics the shape of the square and looks like a rounded square. The dependence
of the diagonal and axial radii on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the deposited liquid for the range of
capillary number. One can see the asymmetry between diagonal ($r_d$) and axial $r_h$ radii for the
capillary
number $Ca\leq\hat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The transition capillary number ($\widehat{Ca}$) between non-axisymmetric case and symmetric is
indicated in a number of
works as $\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space}, $\widehat{Ca}=0.033$ \cite{heil-threedim}. If the capillary number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the droplet dependant on the capillary number. One example for the bubble radii
dependence on the capillary number is in Fig. \ref{fig:heil:three:dim}.

@@

circular shaped capillaries. They also found an empirical correlation which allows one to place

@@

The lattice Boltzmann method (LBM) has emerged as a successful method to simulate

@@

\section{Numerical benchmark}
To be able to properly simulate and compare simulation results with data
published in the literature, one needs to design a numerical benchmark addressing
all the challenges for the lattice Boltzmann method. That includes the diffuse nature of the
interface which needs to have a negligible effect on the results. Also, as far as the density is
uniform, the binary liquid model can address only different viscosities of gas and liquid.
Another challenge is boundary conditions because one needs to couple inlet and outlet
boundary conditions \cite{heil-threedim} for the film thickness to establish.  The geometry of
the problem is also of concern. In the literature, most numerical simulations are done
for a channel with
circular cross section,
which is quite difficult to address in terms of lattice Boltzmann framework. We thoroughly discuss
all the
challenges associated with the lattice Boltzmann microchannel simulations in our recent work
\cite{kuzmin-binary2d}.

@@

Based on these results one can conclude
that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.07$.
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$. $\delta$ is scaled on $H_{\mathrm{eff}}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number we aim for is $\Ca=0.05$. The capillary number obtained from simulations
$\Ca=0.07$. 
\label{fig:grid:profiles}}
\end{figure}
\begin{table}
\begin{tabularx}{0.5\textwidth}{|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$\\
\hline
$1500$&$102$&$0.0694$&$0.0041$&$0.824$&$200000$\\
\hline
$1875$&$127$&$0.0688$&$0.0041$&$0.646$&$250000$\\
\hline
$2250$&$152$&$0.0676$&$0.0040$&$0.539$&$300000$\\
\hline
$2625$&$177$&$0.0679$&$0.0040$&$0.453$&$350000$\\
\hline
$3000$&$202$&$0.0668$&$0.0041$&$0.400$&$400000$\\
\hline
$3375$&$227$&$0.0663$&$0.0039$&$0.355$&$450000$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the interface velocity of the bubble center
measured at the front meniscus. $N_{\mathrm{iter}}$ is the number of simulation time iterations. 
$5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta
(N_y-2)$ is the size of the film in
lattice Boltzmann units.
\label{table:parameters:grid:refinement}}
\end{table}

@@

\section{Conclusion}
This work presents numerical studies of the
Bretherton/Taylor problem using the binary liquid lattice Boltzmann method. The bubble was chosen
sufficiently long for the film
thickness to
stabilize, and periodic boundary conditions were used to keep the simulations robust. The
distance between bubbles was taken large enough to minimize
the mutual influence of neighboring bubbles. The computational
results in terms of capillary number dependence and shape of the bubbles show consistency with the
previously published data in two and three dimensions. The simulations show that in the range of low
capillary numbers the binary liquid model can be used to
simulate gas finger/bubbles propagation in the microchannel. The model is able to catch not only
the film thicknesses but also the change in velocity patterns. 
 An examination of the influence of grid resolution on the
results allowed
us to determine that the phase interface should be resolved as at least $50$ percent of the film
thickness
in order for the simulations to be grid independent. Though our results are specific to the binary
liquid lattice
Boltzmann method, the numerical hints and procedures can be used for any
continuous interface method.
